\bf Model Visualization Techniques for a Social Network Model

### 2017-08-21 paper.Rnw
add adjmat animation

\documentclass[12pt]{article}
% \usepackage{geometry}                % See geometry.pdf to learn the layout options. There are lots.
% \geometry{letterpaper}                   % ... or a4paper or a5paper or ...
%\usepackage{graphicx}
\usepackage[font=small,skip=5pt]{caption}
\usepackage{subcaption}
\usepackage{afterpage}
\usepackage{amssymb}
\usepackage{natbib}
\usepackage{amsmath}
\usepackage{amsfonts}
% \usepackage{color}
\usepackage{multirow}
\usepackage{rotating}
\usepackage[dvipsnames,svgnames,table]{xcolor}
\usepackage{hyperref}
\graphicspath{{figure/}}
% \usepackage{endfloat} % Figures to the end of the document

@@

ThemeNoNet <- theme_bw() %+replace% 
            theme(plot.title = element_text(size = rel(2),
                                            face = 'bold', angle = 0), 
                  axis.title.x = element_text(size = rel(2),
                                            face = 'plain', angle = 0),
                  axis.title.y = element_text(size = rel(2),
                                            face = 'plain', angle = 90),
                  axis.text.x.top = element_text(size = rel(1.5),
                                            face = 'plain', angle = 0), 
                  axis.text.x.bottom = element_text(size = rel(1.5),
                                            face = 'plain', angle = 0), 
                  axis.text.y.left = element_text(size = rel(1.5),
                                            face = 'plain', angle = 0),
                  axis.text.y.right = element_text(size = rel(1.5),
                                            face = 'plain', angle = 0),
                  strip.text.x = element_text(size = rel(2),
                                            face = 'plain', angle = 0, 
                                            margin = margin(t = 3, r = 0, b = 3, l = 0, unit = "pt")),
                  strip.text.y =element_text(size = rel(2),
                                            face = 'plain', angle = 90,margin = margin(t = 3, r = 3, b = 3, l = 3, unit = "pt")),
                  strip.background = element_rect(colour = "black", fill = "white")
                    )
ThemeNet <- theme_net() %+replace% 
            theme(plot.title = element_text(size = rel(2),
                                            face = 'bold', angle = 0), 
                  strip.text.x = element_text(size = rel(2),
                                            face = 'plain', angle = 0, 
                                            margin = margin(t = 3, r = 0, b = 3, l = 0, unit = "pt")),
                  strip.text.y =element_text(size = rel(2),
                                            face = 'plain', angle = 0,margin = margin(t = 3, r = 0, b = 3, l = 0, unit = "pt")),
                  strip.background = element_rect(colour = "black", fill = "white"))
@

@@

<<gof2, fig.cap="An example of viewing the ">>=
qplot(x = log(gof1$`Period 1`$SimulatedTestStat), binwidth = .2) + 
  geom_vline(xintercept = log(gof1$`Period 1`$ObservedTestStat)) + 
  ThemeNoNet + 
  labs(x = "log(Statistic Value)", title = "Distribution of log Mahalanobis Distance", subtitle = "Data shown as vertical line")

@@

The model fitting process in \texttt{RSiena} involves several simulation steps that are hidden from the user. This is practical and efficient if a researcher is primarily interested in fitting one model to a set of longitudinal network data and obtaining parameter estimates. We are more interested in how the models are fit, so we extracted and explored the different steps of that process.

A key component of each step of the SIENA method of moments algorithm is the ``microstep." This step simulates from the model in its current state, $x(t_m)$ with current parameter values $\theta_{0}$, to the next state, $x(t_{m+1})$. This microstep process stops when the simulated network has acheived the same number of differences, $C$, from $x(t_m)$ as $x(t_{m+1})$, where 
$$C = \sum_{i \neq j} |x_{ij}(t_{m+1}) - x_{ij}(t_{m})|.$$ 
This simulation process follows the steps of the continuous-time Markov chain. First, an ``ego node'' is selected to make a change, and then it randomly makes one change in its ties according to the probability, $p_{ij}$, determined by its objective function. The options for change are (1) removing a current tie, (2) adding a new tie, or (3) making no change at all. Saving all of these steps is not efficient if you're only interested in estimating parameter values, but they can be saved and extracted using options in \texttt{RSiena}, which is exactly what we did to create our visualizations. Between two network observations $x(t_m)$ and $x(t_{m+1})$, there can be dozens, hundreds, or even thousands of microsteps, depending on the size of the network. We wanted to see these in-between steps in order to better understand the behavior of the underlying continuous-time Markov chain.

The first vizualization we present here is an animation of microsteps simulated to emulate the transition steps from wave 1 to wave 2 the subset of the friends data shown in Figure ~\ref{fig:alldataviz}. Movies similar to our animation were used to visualize the changes of dynamic networks in \citet{dynnetviz}. When each ego node is selected in a microstep, it is emphasized in the animation, then the associated edge either appears or disappears. If there are no changes at a particular microstep, no changes are seen. Some frames of the animations are shown in Figure~\ref{fig:makinggif}, and the full movie can be viewed at INSERT LINK TO GIF.

@@

The top row of Figure~\ref{fig:makinggif} shows an edge being removed, and the bottom row shows one being added. In both cases, the ego nodes chosen to act change color from black to red, and they also increase greatly in size. In the case of an edge being removed, the edge that currently exists is emphasized with the color and size change that the node also gets, and as the animation proceeds the edge shrinks to disappearing, while the node shrinks and changes color back to the original black. If the edge is being added, it appears colored red from nothing and grows to a large size, then changes color to match the rest of the edges, while the node shrinks and changes color to match the other nodes.

We also use animation to view the changing structure of the adjacency matrix the microsteps. The adjacency matrices for the three waves of friendship data as shown in Figure~\ref{fig:alldataviz} are ordered by node id. There are 16 nodes in the data, numbered 1-16, and that order is used on the $x$ and $y$ axes for the matrix visualization. Viewing the adjacency matrices with this arbitrary ordering does not provide much information to the viewer about the underlying structure of the network. This lack of perceived structure would be exacerbated in an animation. So, we adjust the ordering so that the viewer can perceive more of the structure of the graph. This process is known as matrix seriation \citep{seriation}.

To reorder the vertices for the matrix visualization, we first constructed a cumulative adjaceny matrix, $\mathbf{A}^{cum}$, for the series of microsteps simulating the network from $x(t_m)$ to $x(t_{m+1})$. A single entry in the cumulative adjacency matrix, $\mathbf{A}^{cum}_{ij}$, is the total number of times the edge from node $i$ to node $j$ appears in the network from the initial observation, $x(t_m) \equiv X(0)$ to the final reult of the last microstep, $X(C)$: 
$$ \mathbf{A}^{cum}_{ij} = \sum_{c = 0}^C X_{ij}(c). $$
We then performed a principal component analysis on $\mathbf{A}^{cum}$, and used the values of the first principal component to order the vertices on the $x$ and $y$ axes for the adjacency matrix animation. For one such series of microsteps simulated by \texttt{RSiena}, we present the adjacency matrix ordered by the (arbitrary) vertex id alongside the seriated adjacency microsteps using the first principal component loading on the cumulative adjacency matrix, $\mathbf{A}^{cum}$, in Figure~\ref{fig:adjmatorder}.

@@

pcaordersf$from <- factor(pcaordersf$from, levels = order1)
pcaordersf$to <- factor(pcaordersf$to, levels = rev(order1))

p2 <- ggplot(data = pcaordersf, aes(x = from, y = to)) + geom_tile(color = 'black', aes(fill = as.factor(value))) + 
  scale_fill_manual(values = c('white', 'black')) +
  scale_x_discrete(position = "top", labels = str_replace(order1, "V", "")) + 
  scale_y_discrete(labels = rev(str_replace(order1, "V", ""))) + 
  ThemeNoNet + 
  theme(legend.position = 'none') +
  coord_fixed()

@@

>>>>

\documentclass[12pt]{article}
% \usepackage{geometry}                % See geometry.pdf to learn the layout options. There are lots.
% \geometry{letterpaper}                   % ... or a4paper or a5paper or ...
%\usepackage{graphicx}
\usepackage[font=small,skip=5pt]{caption}
\usepackage{subcaption}
\usepackage{animate}
\usepackage{afterpage}
\usepackage{amssymb}
\usepackage{natbib}
\usepackage{amsmath}
\usepackage{amsfonts}
% \usepackage{color}
\usepackage{multirow}
\usepackage{rotating}
\usepackage[dvipsnames,svgnames,table]{xcolor}
\usepackage{hyperref}
\graphicspath{{figure/}}
% \usepackage{endfloat} % Figures to the end of the document

@@

ThemeNoNet <- theme_bw() %+replace% 
            theme(plot.title = element_text(size = rel(2),
                                            face = 'plain', angle = 0), 
                  axis.title.x = element_text(size = rel(2),
                                            face = 'plain', angle = 0),
                  axis.title.y = element_text(size = rel(2),
                                            face = 'plain', angle = 90),
                  axis.text.x.top = element_text(size = rel(1.5),
                                            face = 'plain', angle = 0), 
                  axis.text.x.bottom = element_text(size = rel(1.5),
                                            face = 'plain', angle = 0), 
                  axis.text.y.left = element_text(size = rel(1.5),
                                            face = 'plain', angle = 0),
                  axis.text.y.right = element_text(size = rel(1.5),
                                            face = 'plain', angle = 0),
                  strip.text.x = element_text(size = rel(2),
                                            face = 'plain', angle = 0, 
                                            margin = margin(t = 3, r = 0, b = 3, l = 0, unit = "pt")),
                  strip.text.y =element_text(size = rel(2),
                                            face = 'plain', angle = 90,margin = margin(t = 3, r = 3, b = 3, l = 3, unit = "pt")),
                  strip.background = element_rect(colour = "black", fill = "white")
                    )
ThemeNet <- theme_net() %+replace% 
            theme(plot.title = element_text(size = rel(2),
                                            face = 'plain', angle = 0), 
                  strip.text.x = element_text(size = rel(2),
                                            face = 'plain', angle = 0, 
                                            margin = margin(t = 3, r = 0, b = 3, l = 0, unit = "pt")),
                  strip.text.y =element_text(size = rel(2),
                                            face = 'plain', angle = 0,margin = margin(t = 3, r = 0, b = 3, l = 0, unit = "pt")),
                  strip.background = element_rect(colour = "black", fill = "white"))
@

@@

<<gof2, fig.cap="An example of viewing the distribution of a statistic of interest from a SAOM. Here, the Mahalanobis distance of the cumulative outdegree values computed on a simulated network to the mean value simulated is shown. The distance computed from the data is shown as a vertical line.", fig.height=3>>=
qplot(x = log(gof1$`Period 1`$SimulatedTestStat), binwidth = .2) + 
  geom_vline(xintercept = log(gof1$`Period 1`$ObservedTestStat)) + 
  ThemeNoNet + 
  labs(x = "log(Statistic Value)", title = "Distribution of log Mahalanobis Distance", subtitle = "Data shown as vertical line")

@@

\st{The last principle of model visualization is to explore the process of fitting the model, instead of just focusing on the end result. This principle is perhaps the most important for SAOMs because} the model fitting process in \texttt{RSiena} involves several simulation steps that are hidden from the user. Hiding the MCMC steps is practical and efficient if a researcher is primarily interested in fitting one model to a set of longitudinal network data, obtaining parameter estimates, \st{and drawing conclusions or making predictions}. We are more interested in \textit{how} the models are fit, so we extracted and explored the different steps of that process.

A key component of each step of the SIENA method of moments algorithm is the ``microstep" process. A series of microsteps is obtained by simulating from the model in its current state, $x(t_m)$ with current parameter values $\mathbf{\theta}_{0} = \{\alpha_{1_0}, \dots \alpha_{{m-1}_0}, \beta_{1_0}, \dots, \beta_{K_0}\}$, to the next state, $x(t_{m+1})$. This microstep process stops when the simulated network has acheived the same number of differences, $C$, from $x(t_m)$ as $x(t_{m+1})$, where 
$$C = \sum_{i \neq j} |x_{ij}(t_{m+1}) - x_{ij}(t_{m})|.$$ 
This simulation process follows the steps of the continuous-time Markov chain. \st{Each tie change in the CTMC is referred to as one ``microstep". At each microstep}, an ``ego node" is selected to make a change, and the chosen ego node randomly makes one change in its ties according to the probabilities, $\{p_{ij}: i\neq j \in \{1, \dots, n\}\}$, determined by its objective function. The options for change are (1) removing a current tie, (2) adding a new tie, or (3) making no change at all. Saving and exploring all of these steps is not computationally efficient if one is only interested in estimating parameter values, but they can be saved and extracted using options in \texttt{RSiena}, which is exactly what we did to create our visualizations. Between two network observations $x(t_m)$ and $x(t_{m+1})$, there can be dozens, hundreds, or even thousands of microsteps, depending on the size of the network and the number of changes between two network observations. We wanted to view these in-between steps in order to better understand the behavior of the underlying continuous-time Markov chain.

The first vizualization we present here is an animation of microsteps simulated to emulate the transition steps of the CTMC from wave 1 to wave 2 of the small friendship network example shown in Figure ~\ref{fig:alldataviz}. Movies similar to this animation were used to visualize the changes of dynamic networks in \citet{dynnetviz}. When each ego node is selected in a microstep, it is emphasized in the animation, then the associated edge either appears or disappears. If there are no changes at a particular microstep, no changes are seen. Some frames of the animations are shown in Figure~\ref{fig:makinggif}, and the full movie can be viewed at INSERT LINK TO GIF.

@@

The top row of Figure~\ref{fig:makinggif} shows an edge being removed, and the bottom row shows one being added. In both cases, the ego nodes chosen to act change color from black to red, and they also increase greatly in size. In the case of an edge being removed, shown in the top row, the edge that currently exists is emphasized with the same color and size change that the node gets, and as the animation proceeds the edge shrinks to nothing, as the ego node shrinks and changes color back to its original black. If an edge is being added, as in the bottom row of the figure, the ego node's appearances changes in the sampe way as when the edge is being removed, whil the edge appears colored red from nothing, and grows to a large size, then changes color and size to match the rest of the edges, while the node shrinks and changes color to match the other nodes.

We also use animation to view the changing structure of the adjacency matrix the microsteps. The adjacency matrices for the three waves of friendship data as shown in Figure~\ref{fig:alldataviz} are ordered by node id. There are 16 nodes in the data, numbered 1-16, and that order is used on the $x$ and $y$ axes for the matrix visualization. Viewing the adjacency matrices with this arbitrary ordering does not provide much information to the viewer about the underlying structure of the network. This lack of perceived structure would be exacerbated in an animation, so we adjust the ordering so that the viewer can perceive more of the structure of the graph. This process is known as matrix seriation \citep{seriation}.

@@

pcaordersf$from <- factor(pcaordersf$from, levels = rev(order1))
pcaordersf$to <- factor(pcaordersf$to, levels = order1)

p2 <- ggplot(data = pcaordersf, aes(x = to, y = from)) + geom_tile(color = 'black', aes(fill = as.factor(value))) + 
  scale_fill_manual(values = c('white', 'black')) +
  scale_x_discrete(position = "top", labels = str_replace(order1, "V", "")) + 
  scale_y_discrete(labels = rev(str_replace(order1, "V", ""))) + 
  ThemeNoNet + 
  theme(legend.position = 'none') +
  coord_fixed()

@@

To reorder the vertices for the matrix visualization, we first constructed a cumulative adjaceny matrix, $\mathbf{A}^{cum}$, for the series of microsteps simulating the network from $x(t_m)$ to $x(t_{m+1})$. A single entry in the cumulative adjacency matrix, $\mathbf{A}^{cum}_{ij}$, is the total number of times the edge from node $i$ to node $j$ appears in the network from the initial observation, $x(t_m) \equiv X(0)$ to the final result of the last microstep, $X(R)$, where $R$ is the total number of microsteps taken: 
$$ \mathbf{A}^{cum}_{ij} = \sum_{r = 0}^R X_{ij}(r). $$
We then performed a principal component analysis on $\mathbf{A}^{cum}$, and used the values of the first principal component to order the vertices on the $x$ and $y$ axes for the adjacency matrix animation. For one such series of microsteps simulated by \texttt{RSiena}, we present the adjacency matrix ordered by the (arbitrary) vertex id alongside the seriated adjacency microsteps using the first principal component loading on the cumulative adjacency matrix, $\mathbf{A}^{cum}$, in Figure~\ref{fig:adjmatorder}.

<<adjmatani, fig.cap='The adjacency matrix visualization animation for one series of microsteps. At the beginning of the animation, there are two connected components. By the end, the components appear to be more spread out.', fig.show='animate', warning = FALSE, message = FALSE, interval=.2, results='hide', fig.height=4>>=
adjmatall <- matrix(0, nrow = 16, ncol = 16)
rownames(adjmatall) <- paste0("V", 1:16)
colnames(adjmatall) <- paste0("V", 1:16)
dat2 <- sfmsall %>% filter(rep == 1)
adjmatall2 <- NULL
for(i in 0:55){
  MS <- dat2[dat2$ms == i,c("from", "to")]
  myadjmat <- adjmatall
  for(j in 1:nrow(MS)){
    myadjmat[MS[j,"from"][[1]],MS[j,"to"][[1]]] <- 1
  }
  adjmatall2 <- rbind(adjmatall2, myadjmat)
}

adjmatall2 <- cbind(adjmatall2, rep(0:55, each = 16))
adjmatall2 <- data.frame(adjmatall2)
adjmatall2$from <- rep(paste0("V", 1:16), 56)
names(adjmatall2)[17] <- "ms"
adjmatall2 %>% gather(to, val, V1:V16) -> adjmatall3

adjmatall3$to <- factor(adjmatall3$to, levels = order1)
adjmatall3$from <- factor(adjmatall3$from, levels = rev(order1))
plots <- list()
for(i in 0:55){
plots[[i + 1]] <- ggplot(data = adjmatall3 %>% filter(ms == i)) + 
  geom_tile(aes(x = to, y = from, fill = as.factor(val)), color = 'black') + 
  scale_fill_manual(values = c("white", "black")) + 
  scale_x_discrete(position = "top", labels = str_replace(order1, "V", "")) + 
  scale_y_discrete(labels = rev(str_replace(order1, "V", ""))) + 
  ThemeNoNet + 
  theme(legend.position = 'none') +
  coord_fixed() 
}
plots 
@
